Local simple connectedness of boundaries of hyperbolic groups

### 2020-04-02 paper.tex
Proof read new section 3.2
Proof read sections 3 & 4

\begin{lemma}\label{lem:nonempty_intersection}
  Every bi-infinite geodesic with one end point equal to $\xi$ meets $H$.
  Furthermore, the intersection between the geodesic and $H$ is contained in a
  ball with centre $v$ and radius depending only on $\delta$ and the distance
  from the $v$ to the geodesic.
\end{lemma}

@@

  To see that it is bounded, let $p \in H \intersection \alpha$. Since $z_i \to
  \xi$ as $i\to \infty$, $z_i$ is between $\xi$ and $p$ on $\alpha$ for $i$
  sufficiently large. Fix $i \geq i_0$ so that this is true and also so that $p
  \in H_i$, so $\dist(g_i, p) = n_i$. Then the point $p$ has the properties
  assumed of $p_i$, so the argument above shows that $\dist(p, v) \leq
  2\dist(x,v) + 2\delta$. Therefore the intersection $H \intersection \alpha$
  is bounded.
\end{proof}

@@

\subsection{Projections to the horosphere}

\begin{definition}
  For each $i$, let $U_i \subset \boundary G$ be the set $\left(g_i \composed
  p^\infty_{n_i} \composed g_i^{-1}\right)^{-1} H_i$ and let $q_i \from U_i \to
  H_i$ be the restriction of $g_i \composed p^\infty_{n_i}\composed g_i^{-1}$
  to this set.
\end{definition}

@@

\begin{lemma}\label{lem:describing_q_i}
  Let $\zeta \in \boundary G$ be a point such that
  \begin{equation*}
    N_{2\delta+1}^{g_i\cdot S_{n_i}} \set{\eta(n_i)\suchthat\text{$\eta$ a geodesic ray with
        $\eta(0) = g_i$ and $\eta(\infty) = \zeta$}} \subset H_i,
  \end{equation*}
  where $N_{2\delta+1}^{g_i\cdot S_{n_i}}$ denotes a $2\delta+1$-neighbourhood
  in $g_i\cdot S_{n_i}$ taken with respect to the metric $\dist$.
  Then $\zeta \in U_i$ and $q_i(\zeta)$ is contained in the simplex of $H_i$
  spanned by this set of vertices.
\end{lemma}

@@

\begin{proof}
  Suppose that a compact subset $C$ of $\boundary G - \set\xi$ is not contained
  in $U_i$ for infinitely many $i$. Then after passing to a subsequence there
  exists for each $i$ a point $\zeta_i \in C\setminus U_i$, so by
  Lemma~\ref{lem:describing_q_i} for each $i$ there is a geodesic ray
  $\alpha_i$ from $g_i$ to $\zeta_i$ that meets $N_{2\delta+1}^{g_i\cdot
  S_{n_i}}(g_i\cdot S_{n_i} - H_i)$, and so does not meet $H_i\intersection
  N_{i-4\delta-1}(v)$, which is equal to $H \intersection N_{i-4\delta-1}(v)$.
  After passing to a subsequence, the rays $\alpha_i$ converge uniformly on
  compact subsets to a bi-infinite geodesic from $\xi$ to some point $\zeta \in
  C$ that does not meet $H$, contradicting
  Lemma~\ref{lem:nonempty_intersection}.
\end{proof}

@@

\begin{proof}
  Let $B\subset H$ be a bounded set containing all points where geodesics with
  one limit point equal to $\xi$ and the other in $C$ meet $H$; this set exists
  by \cref{lem:bounded_intersection}.

  Fix $\alpha$ to be some geodesic with $\alpha(-\infty)=\xi$ and
  $\alpha(\infty)\in C$ and let $x$ be a point of intersection between $\alpha$
  and $B$. Since the triangle with vertices $x$, $g_i$ and $\xi$ is
  $\delta$-slim, and $\dist(x, [\xi, g_i]) \to \infty$ as $i\to\infty$, for
  each $i$ there exists a point $y_i$ on $[g_i, x]$ with $\dist(y_i, \alpha)
  \leq \delta$ and so that $y_i \to \xi$ as $i\to\infty$. Let $z_i \in \alpha$
  be such that $\dist(y_i, z_i)\leq\delta$.

@@

  Let $V$ be the following set. It is easy to check that this is a
  neighbourhood of $\zeta$.
  \begin{equation*}
    V = \set{\beta(\infty) \suchthat \text{$\beta$ a geodesic with
            $\beta(-\infty) = \xi$ and $\dist(\beta, x) \leq 3\delta+1$}}
  \end{equation*}
  The neighbourhood $V$ is relatively compact in $\boundary G - \set\xi$ so
  we may take $i_0$ such that $V \subset U_i$ for $i \geq i_0$. Enlarge $i_0$
  so that for any $i \geq i_0$, $\dist([\xi, g_i], x) \geq 4\delta+1$.

@@

We begin this section by recalling work of Bestvina and Mess, Bowditch, and
Swarup, which characterise the hyperbolic groups with locally path connected
boundary. In~\cite{bestvinamess91}, Bestvina and Mess introduce the following
condition that a hyperbolic group might satisfy. (Regarding the constants,
note that with our convention that ideal triangles are $\delta$-hyperbolic,
$C$ in the cited paper can be taken to be equal to $\delta$.)

@@

\subsection{An modified condition}

In this section we introduce a related condition. This condition is stronger
than $\ddag_M$, but, using Bestvina and Mess's result, we show that it holds
under the same assumption. This condition reinterprets the paths in $\ddag_M$
as edge paths in the complexes $K_n$, with the additional assumption that
these paths approximate paths that factor through the projection map
$p^\infty_n$. This interpretation strengthens the link between
path-connectedness of the boundary and connectedness of the complexes $K_n$.

\begin{definition}
  We say that $G$ satisfies $\ddag'_M$ if there exist numbers $L$ and $n_0$
  such that, for any $n \geq n_0$, any $\zeta_1$ and $\zeta_2$ in $\boundary
  G$ such that $\newdist(p^\infty_n(\zeta_1), p^\infty_n(\zeta_2)) \leq M$,
  and any vertices $x_1$ and $x_2$ of the minimal simplices of $K_n$
  containing $p^\infty_n(\zeta_1)$ and $p^\infty_n(\zeta_2)$, there is a path
  $\alpha\from I\to \boundary G$ from $\zeta_1$ to $\zeta_2$ in $\boundary G$
  such that $p^\infty_n\composed\alpha$ admits a simplicial approximation
  connecting $x_1$ to $x_2$ in $K_n$ after at most $L$ barycentric
  subdivisions of $I$.
\end{definition}

@@

\begin{proof}
  Suppose $\ddag'_M$ does not hold. Then for any $i$ there exists a number $n_i
  \geq i$, points $\zeta_1^i$ and $\zeta_2^i$ in $\boundary G$ and vertices
  $x_1^i$ and $y_1^i$ of the minimal simplex of $K_{n_i}$ containing 
  $p^\infty_{n_i}(\zeta^i_1)$ and $p^\infty_{n_i}(\zeta^i_2)$ such that there
  is no path $\alpha$ from $\zeta_1^i$ to $\zeta_2^i$ in $\boundary G$ where
  $p^\infty_{n_i}\composed\alpha$ admits a simplicial approximation after at
  most $i$ barycentric subdivisions of the interval.

@@

  Since $\boundary G$ is locally path connected, for $j$ equal to $1$ or $2$
  there exists a neighbourhood of $\zeta_j$ such that any point in this
  neighbourhood is joined to $\zeta_j$ by a path contained in the
  $\epsilon$-neighbourhood of $\zeta_j$. Now increase $i_0$ to ensure that
  $\zeta_1^i$ and $\zeta_2^i$ are contained in these neighbourhoods for $i \geq
  i_0$. Let $\alpha_1^j$ be a path in the ball of radius $\epsilon$ centred at
  $\zeta_j$ connecting $\zeta_j^i$ to $\zeta_j$.

  Then the concatenation $\alpha^i\from[0,1]\to \boundary G$ of $\alpha_1^i$,
  $\beta$ and $\alpha_2^i$ gives a path in $C$ from $\zeta_1^i$ to $\zeta_2^i$
  such that $I$ can be covered by boundedly many open intervals, each of which
  has image under $\alpha^i$ with diameter at most $\epsilon$. It follows that
  the composition of $\alpha^i$ with $q_i$ is a path that admits a simplicial
  approximation connecting $x_1$ to $x_2$ after boundedly many subdivisions of
  the interval. For $i \geq i_0$ large enough this gives a contradiction.
\end{proof}

@@

  For an edge $e$ of $K_n$ connecting vertices $x_1$ and $x_2$,
  $\dist(i^n_{n+1}(x_1), i^n_{n+1}(x_2))$ is at most $2(2\delta + 1) + D = D +
  4\delta + 2$. It follows from $\ddag'_{D+4\delta+2}$ that there is a path
  $\alpha$ from $\gamma_{x_1}(\infty)$ to $\gamma_{x_2}(\infty)$ in $\boundary
  G$ so that $p^\infty_{n+1}\composed\alpha$ admits a simplicial approximation
  connecting $i^n_{n+1}(x_1)$ to $i^n_{n+1}(x_2)$ after $L$ subdivisions.
  Define $i^n_{n+1}$ to be this simplicial map on $\sd^L e$.

@@

\subsection{A two-dimensional version}

@@

\begin{definition}
  Let $\Delta$ be the standard 2-simplex.  We say that $G$ satisfies $\S_M$ if
  there exist $L$ and $n_0$ such that for any $n\geq n_0$ and any simplicial
  map $f\from \sd^M\boundary\Delta \to K_n$ that is a simplicial approximation
  to a map that factors through $p^\infty_n \from \boundary G \to K_n$, $f$
  admits an extension to $\Delta$ that is affine on $\sd^L \Delta$. (We say
  that a map between simplicial complexes is affine if its restriction to any
  simplex of the domain is affine.)
\end{definition}

@@

\begin{proof}
  This follows from the main theorem of~\cite{zeeman64}: note that the degree
  $L$ subdivision of a complex can be obtained from any degree $L$ relative
  subdivision by further subdividing some simplices, so any map that is affine
  on the relative subdivision is also affine on the full subdivision.
\end{proof}

>>>>

\begin{lemma}\label{lem:nonempty_intersection}
  Every bi-infinite geodesic with one end point equal to $\xi$ meets $H$.

  Furthermore, the intersection between the geodesic and $H$ is contained in a
  ball with centre $v$ and radius depending only on $\delta$ and the distance
  from the $v$ to the geodesic.
\end{lemma}

@@

  To see that it is bounded, let $p \in H \intersection \alpha$. Since $z_i \to
  \xi$ as $i\to \infty$, $z_i$ is between $\xi$ and $p$ on $\alpha$ for $i$
  sufficiently large. Fix $i \geq i_0$ so that this is true and also so that $p
  \in H_i$, so $\dist(g_i, p) = n_i$. Then the point $p$ has the properties
  assumed of $p_i$ above, so the argument above shows that $\dist(p, v) \leq
  2\dist(x,v) + 2\delta$. Therefore the intersection $H \intersection \alpha$
  is bounded.
\end{proof}

@@

\subsection{Projections from the boundary to the horosphere}

\begin{definition}
  For each $i$, let $U_i \subset \boundary G$ be the set $\left(g_i \composed
  p^\infty_{n_i} \composed g_i^{-1}\right)^{-1} H_i$ and let $q_i \from U_i \to
  H_i$ be the restriction of $g_i \composed p^\infty_{n_i}\composed g_i^{-1}$
  to $U_i$.
\end{definition}

@@

\begin{lemma}\label{lem:describing_q_i}
  Let $\zeta \in \boundary G$ be a point such that
  \begin{equation*}
    N_{2\delta+1}^{g_i\cdot S_{n_i}} \set{\eta(n_i)\suchthat\text{$\eta$ a geodesic ray with
        $\eta(0) = g_i$ and $\eta(\infty) = \zeta$}}
  \end{equation*}
  is contained in $H_i$, where $N_{2\delta+1}^{g_i\cdot S_{n_i}}$ denotes a
  $2\delta+1$-neighbourhood in $g_i\cdot S_{n_i}$ taken with respect to the
  word metric $\dist$.  Then $\zeta \in U_i$ and $q_i(\zeta)$ is contained in
  the simplex of $H_i$ spanned by this set of vertices.
\end{lemma}

@@

\begin{proof}
  Suppose that a compact subset $C$ of $\boundary G - \set\xi$ is not contained
  in $U_i$ for infinitely many $i$. Then after passing to a subsequence there
  exists for each $i$ a point $\zeta_i \in C\setminus U_i$, so by
  Lemma~\ref{lem:describing_q_i} for each $i$ there is a geodesic ray
  $\alpha_i$ from $g_i$ to $\zeta_i$ that meets $N_{2\delta+1}^{g_i\cdot
  S_{n_i}}(g_i\cdot S_{n_i} - H_i)$, and so does not meet $H_i\intersection
  N_{i-4\delta-1}(v)$, which is equal to $H \intersection N_{i-4\delta-1}(v)$.
  After passing to a subsequence and reparametrising, the rays $\alpha_i$
  converge uniformly on compact subsets to a bi-infinite geodesic from $\xi$ to
  some point $\zeta \in C$ that does not meet $H$, contradicting
  Lemma~\ref{lem:nonempty_intersection}.
\end{proof}

@@

\begin{proof}
  Let $B\subset H$ be the set of points of intersection between $H$ and
  geodesics with one end point equal to $\xi$ and the other in $C$; this set is
  bounded by Lemma~\ref{lem:bounded_intersection}.

  Fix $\alpha$ to be some geodesic with $\alpha(-\infty)=\xi$ and
  $\alpha(\infty)\in C$ and let $x$ be a point of intersection between $\alpha$
  and $H$, so $x \in B$.  Since the triangle with vertices $x$, $g_i$ and $\xi$
  is $\delta$-slim, and $\dist(x, [\xi, g_i]) \to \infty$ as $i\to\infty$, for
  each $i$ there exists a point $y_i$ on $[g_i, x]$ with $\dist(y_i, \alpha)
  \leq \delta$ and so that $y_i \to \xi$ as $i\to\infty$. Let $z_i \in \alpha$
  be such that $\dist(y_i, z_i)\leq\delta$.

@@

  Let $V$ be the following neighbourhood of $\zeta$.
  \begin{equation*}
    V = \set{\beta(\infty) \suchthat \text{$\beta$ a geodesic with
            $\beta(-\infty) = \xi$ and $\dist(\beta, x) \leq 3\delta+1$}}.
  \end{equation*}
  The neighbourhood $V$ is relatively compact in $\boundary G - \set\xi$ so
  we may take $i_0$ such that $V \subset U_i$ for $i \geq i_0$. Enlarge $i_0$
  so that for any $i \geq i_0$, $\dist([\xi, g_i], x) \geq 4\delta+1$.

@@

We begin this section by recalling work of Bestvina and Mess, Bowditch, and
Swarup, which characterise the hyperbolic groups with locally path connected
boundary. In~\cite{bestvinamess91}, Bestvina and Mess introduce the following
condition that a hyperbolic group might satisfy. (When comparing the constants
appearing in our definition of the condition with that in the cited paper, note
that with our convention that ideal triangles are $\delta$-hyperbolic, Bestvina
and Mess's constant $C$ can be taken to be equal to $\delta$.)

@@

\subsection{A modified condition}

In this section we introduce a related condition. This condition appears
stronger than $\ddag_M$, but, using Bestvina and Mess's result, we show that it
holds under the same assumption. This condition reinterprets the paths in
$\ddag_M$ as edge paths in the complexes $K_n$, with the additional assumption
that these paths approximate paths that factor through the projection map
$p^\infty_n$. This interpretation strengthens the link between
path-connectedness of the boundary and connectedness of the complexes $K_n$.

\begin{definition}
  Let $I$ be the interval $[0,1]$, which we identify with a 1-simplex.

  We say that $G$ satisfies $\ddag'_M$ if there exist numbers $L$ and $n_0$
  such that, for any $n \geq n_0$, any $\zeta_1$ and $\zeta_2$ in $\boundary G$
  such that $\newdist(p^\infty_n(\zeta_1), p^\infty_n(\zeta_2)) \leq M$, and
  any vertices $x_1$ and $x_2$ of simplices of $K_n$ containing
  $p^\infty_n(\zeta_1)$ and $p^\infty_n(\zeta_2)$, there is a path $\alpha\from
  I\to \boundary G$ from $\zeta_1$ to $\zeta_2$ in $\boundary G$ such that
  $p^\infty_n\composed\alpha$ admits a simplicial approximation connecting
  $x_1$ to $x_2$ in $K_n$ after at most $L$ barycentric subdivisions of $I$.
\end{definition}

@@

\begin{proof}
  Suppose $\ddag'_M$ does not hold. Then for any $i$ there exists a number $n_i
  \geq i$, points $\zeta_1^i$ and $\zeta_2^i$ in $\boundary G$ and vertices
  $x_1^i$ and $y_1^i$ of simplices of $K_{n_i}$ containing
  $p^\infty_{n_i}(\zeta^i_1)$ and $p^\infty_{n_i}(\zeta^i_2)$ such that
  $\newdist(p^\infty_{n_i}(\zeta^i_1), p^\infty_{n_i}(\zeta^i_2)) \leq M$ and
  there is no path $\alpha$ from $\zeta_1^i$ to $\zeta_2^i$ in $\boundary G$
  where $p^\infty_{n_i}\composed\alpha$ admits a simplicial approximation after
  at most $i$ barycentric subdivisions of the interval.

@@

  Since $\boundary G$ is locally path connected, for $j$ equal to $1$ or $2$
  there exists a neighbourhood of $\zeta_j$ such that any point in this
  neighbourhood is joined to $\zeta_j$ by a path contained in the
  $\epsilon$-neighbourhood of $\zeta_j$. Now increase $i_0$ to ensure that
  $\zeta_1^i$ and $\zeta_2^i$ are contained in these neighbourhoods for $i \geq
  i_0$. For $j=1$ and $j=2$ let $\alpha_1^j$ be a path in the ball of radius
  $\epsilon$ centred at $\zeta_j$ connecting $\zeta_j^i$ to $\zeta_j$.

  Then the concatenation $\alpha^i\from[0,1]\to \boundary G$ of $\alpha_1^i$,
  $\beta$ and $\alpha_2^i$ gives a path in $C$ from $\zeta_1^i$ to $\zeta_2^i$
  such that $I$ can be covered by boundedly many open intervals, each of which
  has image under $\alpha^i$ with diameter at most $\epsilon$. It follows that
  the composition of $\alpha^i$ with $q_i$ is a path in $H$ that admits a
  simplicial approximation connecting $x_1$ to $x_2$ after boundedly many
  subdivisions of the interval. For $i \geq i_0$ large enough this gives a
  contradiction.
\end{proof}

@@

  For an edge $I$ of $K_n$ connecting vertices $x_1$ and $x_2$,
  $\dist(i^n_{n+1}(x_1), i^n_{n+1}(x_2))$ is at most $2(2\delta + 1) + D = D +
  4\delta + 2$. It follows from $\ddag'_{D+4\delta+2}$ that there is a path
  $\alpha$ from $\gamma_{x_1}(\infty)$ to $\gamma_{x_2}(\infty)$ in $\boundary
  G$ so that $p^\infty_{n+1}\composed\alpha$ admits a simplicial approximation
  connecting $i^n_{n+1}(x_1)$ to $i^n_{n+1}(x_2)$ after $L$ subdivisions.
  Define $i^n_{n+1}$ to be this simplicial map on $\sd^L I$.

@@

\subsection{Extending maps from the boundary of a 2-simplex}

@@

\begin{definition}
  Let $\Delta$ be the standard 2-simplex.

  We say that $G$ satisfies $\S_M$ if there exist $L$ and $n_0$ such that for
  any $n\geq n_0$ and any simplicial map $f\from \sd^M\boundary\Delta \to K_n$
  that is a simplicial approximation to a map that factors through $p^\infty_n
  \from \boundary G \to K_n$, $f$ admits an extension to $\Delta$ that is
  affine on $\sd^L \Delta$. (We say that a map between simplicial complexes is
  affine if its restriction to any simplex of the domain is affine.)
\end{definition}

@@

\begin{proof}
  This follows from the main theorem of~\cite{zeeman64}: note that the degree
  $L$ subdivision of a complex can be obtained from any degree $L$ relative
  subdivision by further subdividing some simplices, so any map that is a
  simplicial map on the relative subdivision is affine on the full subdivision.
\end{proof}
